---
title: "COVID-19 outbreak"
author: 
    Eduard Bakstein,
    eduard.bakstein@nudz.cz,
    http://bakstein.net,
    @edbakstein
date: "`r format(Sys.time(), '%d. %m. %Y')`"
output:
  pdf_document: default
  word_document: default
---

```{r settings-packages, echo=F,warning=F,message=F}

# data handling + plots
library(tidyverse)
library(lubridate)
library(RColorBrewer)
library(knitr)
library(directlabels)

# web scraping
library(rvest)
library(jsonlite)

library(tmap)

# epidemiology
# library(earlyR)
# library(EpiEstim)

# setup
knitr::opts_chunk$set(echo = F, warning = F,message = F,fig.width = 8,fig.height=5,cache=FALSE,dev="pdf")
tmp<-Sys.setlocale("LC_TIME", "C")

figDir<-'../fig/'
dataDir<-'../data/'
dateStr<-format(Sys.time(),'%Y%m%d-%H%M')


vars<-c('cases','incident','deaths')
colSc<-scale_colour_manual(
  values = brewer.pal(3,"Set1")[c(2,3,1)],
  labels = vars,
  aesthetics = c("colour", "fill"),
  name = 'variable'
)
 
# data-loading func
source('covidDataLoad.R')


```


## Intro

The aim is to do exploratory analyses of the COVID-19 cases in selected countries and see how the models evolve as new cases are reported. Can be used as a quick&dirty basis for your own analyses. Special focus given to Czech Rep., as this is my country of origin.

Source code (R markdown): https://github.com/ebakstein/covid19.git


**Resources:**

- https://rviews.rstudio.com/2020/03/05/covid-19-epidemiology-with-r/ (basis of this rmd and analysis)
- https://timchurches.github.io/blog/posts/2020-02-18-analysing-covid-19-2019-ncov-outbreak-data-with-r-part-1/

**Interesting models and remarks **

- https://blog.ephorie.de/epidemiology-how-contagious-is-novel-coronavirus-2019-ncov
- https://arxiv.org/pdf/2002.00418v1.pdf

**Data sources:**

- https://en.wikipedia.org/w/index.php?title=2020_coronavirus_outbreak_in_the_United_States&oldid=944107102
- https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series


## TO DO

 - modelling


## Load data
Using the Jons-Hopkins github repository to obtain current data, see
https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series (updated daily around 23:59 UTC)

```{r load-data}

covid<-covidLoadJHU()
 

```

Obtain data about population in each country by scraping the wikipedia page: 'https://en.wikipedia.org/wiki/List_of_countries_by_population_(United_Nations)'

```{r number-inhabit}

# attach population data from wikipedia
covid <- attachPopCounts(covid)

```


### Plot selected countries
```{r}
selCountries<-c('Czech Republic','Italy','Iran','Germany','Spain','France','Austria','Slovakia','Poland','South Korea','United States','United Kingdom')

# selCountries<-c('Czech Republic','Finland','Austria','Sweden','Denmark','Norway')


# cases in selected countries
covid %>% filter(country %in% selCountries) %>%    ggplot(aes(x=Date,y=cases,col=country)) + geom_line() + ggtitle('cummulative cases - comparison') + scale_x_date(expand=expand_scale(add=10),limits = c(as.Date('2020-02-15'),NA)) + geom_dl(aes(label = country), method = list(dl.combine("last.points"), cex = 0.6)) + theme(legend.position = 'none')
selCountries<- c(selCountries,'China')

```

Note: China number of cases too high to be plot on linear y axis with other countries

```{r}
covid %>% filter(country %in% selCountries) %>%  ggplot(aes(x=Date,y=cases,col=country)) + scale_y_continuous(trans='log10',breaks = 10^(0:5),labels = function(x) format(x, scientific = FALSE)) + geom_line() + ggtitle('cummulative cases - Log plot') + scale_x_date(expand=expand_scale(add=10)) + geom_dl(aes(label = country), method = list(dl.combine("last.points"), cex = 0.6)) + theme(legend.position = 'none')# + facet_wrap(~country) + theme(legend.position = 'none')

```

Note: China added to the log plot


```{r per-capita}

covid %>% filter(country %in% selCountries) %>%  ggplot(aes(x=Date,y=casesPer1M,col=country)) + scale_y_continuous(trans='log10',labels = function(x) format(x, scientific = FALSE))+ scale_x_date(expand=expand_scale(add=10)) + geom_line() + ggtitle('cummulative cases per 1 mil. inhabitants - Log plot') + geom_dl(aes(label = country), method = list(dl.combine("last.points"), cex = 0.6)) + theme(legend.position = 'none')   # + facet_wrap(~country)


```


### Cummulative stats for Individual selected countries
```{r plot-it}


for(cntr in selCountries){
  
  cdata<-covid %>% filter(country==cntr)
  
  outbreakDate<- min(cdata$Date[cdata$cases>0])
  
  plt <-  cdata[,1:5] %>% reshape2::melt(id=c('country','Date')) %>% mutate(variable=fct_relevel(variable,vars)) %>% ggplot(aes(x=Date,y=value,col=variable)) + geom_line() + colSc+ ggtitle(paste0('COVID19 - ',cntr)) + xlim(outbreakDate-days(3),NA)# + facet_wrap(~country) + theme(legend.position = 'none')
  
  
  print(plt)
}

```

### Latest number of cases across countries

According to the Johns-Hopkins data

```{r latest-numbers}

maxCounts <- covid %>% filter(country %in% selCountries) %>% group_by(country) %>% summarize(dateMax=Date[which.max(cases)],count=max(cases), deaths=deaths[which.max(cases)],newCases=incident[which.max(cases)]) %>% mutate(propIncrease=newCases/(count-newCases),deathRate=deaths/count) %>% ungroup()
  
kable(maxCounts)

```



## Plot and analyze Czech Rep.  - CZ data

obtain latest czech data from MZ / apify https://api.apify.com/v2/key-value-stores/K373S4uCFR9W1K8ei/records/LATEST?disableRedirect=true (parsed from https://onemocneni-aktualne.mzcr.cz/covid-19)

```{r cech-apify-json}

# load CZ covid data
tmp<-covidLoadCZ()
covidCZ <- tmp$covidCZ
covidCZregions <- tmp$covidCZregions
covidCZAgeSex <- tmp$infectedByAgeSex

```



### exploratory plots: UZIS data

```{r plot-CZ}
covidCZ %>% mutate(propIncrease=propIncrease*100) %>% reshape2::melt(id=c('date')) %>% ggplot(aes(x=date,y=value,col=variable)) + geom_line() + ggtitle('COVID-19 Czech Republic: number of cases and tests (UZIS data)')


prop<-covidCZ[,c('date','propIncrease','newCases')]  %>% mutate(propIncrease=propIncrease*100) %>% reshape2::melt(id=c('date')) %>% ggplot(aes(x=date,y=value,col=variable)) + geom_line() + scale_x_date(breaks='weeks', limits=c(as.Date('2020-02-26'),NA)) + ggtitle('COVID-19 Czech Republic: proportional increase and new cases (UZIS data)') + theme(legend.position=c(0.02,0.97),legend.justification = c(0,1))

prop

ggsave(paste0(figDir,'CzPropIncrease-',dateStr,'.png'),prop,width=20,height=15,units='cm',scale=.8)


meanIncreaseCZ<-mean(covidCZ$propIncrease,na.rm=TRUE)

meanIncreaseCZl5<-mean(covidCZ$propIncrease[nrow(covidCZ)+(-4:0)],na.rm=TRUE)

outbreakCZ<-min(covidCZ$date[covidCZ$totalCases>0],na.rm=T)

# infected by age and sex
qp<-ggplot(covidCZAgeSex,aes(x=ageGrp,y=infected,fill=sex)) + geom_bar(stat='identity') + ggtitle('infected by afe and sex')

qp

qp + facet_grid(sex~.)

```

first case in CZ: `r outbreakCZ`

Mean daily increase: `r format(100*meanIncreaseCZ,digits=3)`%

The diseased count doubles on average every `r format(1/log2(1+meanIncreaseCZ),digits=3)` days (`r format(24/log2(1+meanIncreaseCZ),digits=3)` hours).

Mean daily increase (last 5 days): `r format(100*meanIncreaseCZl5,digits=3)`%

The diseased count doubles on average every `r format(1/log2(1+meanIncreaseCZl5),digits=3)` days (`r format(24/log2(1+meanIncreaseCZl5),digits=3)` hours).

#### CZ data by region
 credit: Vladimir Mazoch
# ```{r map-cz}
# 
# data <- covidCZregions
# data <- data.frame(matrix(unlist(data), nrow=length(data), byrow=T))
# names(data) <- c("county", "Počet případů")
# data$`Počet případů` <- as.numeric(as.character(data$`Počet případů`))
# data$NAME_1 <- gsub(" kraj", "", data$county)
# data$NAME_1 <- gsub("Hlavní město Praha", "Prague", data$NAME_1)
# 
# # https://gadm.org/download_country_v3.html
# kraje <- readRDS(gzcon(url("https://biogeo.ucdavis.edu/data/gadm3.6/Rsf/gadm36_CZE_1_sf.rds")))
# kraje <- merge(kraje, data)
# 
# # tmap ----
# 
# mapa <- tm_shape(kraje) +
#   tm_fill("totalCases", breaks = c(10, 25, 50, 100, 150, 200, 300)) +
#   tm_borders(col = "gray80") +
#   tm_layout(
#     legend.outside = TRUE,
#     frame = FALSE,
#     main.title = paste("CZ: total COVID-19 cases"),
#     main.title.size = .8
#   )
# 
# mapa

```


#### Increase rate


```{r vs-typical-increases}
# model of growth - starts 2 weeks ago
strtX<-Sys.Date()-days(5);
strtY<-covidCZ$totalCases[covidCZ$date==strtX]

modelX<-data.frame(date=seq(outbreakCZ,Sys.Date()+days(2),by='days'))
modelX<-modelX %>% mutate(daysOutbreak=as.numeric(difftime(modelX$date,strtX,units='days'))) %>%  mutate(dbl2d=strtY*2^(daysOutbreak/2),dbl3d=strtY*2^(daysOutbreak/3),dbl4d=strtY*2^(daysOutbreak/4),dbl5d=strtY*2^(daysOutbreak/5))



modelX %>%reshape2::melt(id.vars='date',measure.vars=grep('^dbl',names(modelX))) %>%
          ggplot(aes(x=date,y=value,col=variable)) +
          geom_line(linetype='dashed') + 
          geom_line(data=covidCZ,aes(x=date,y=totalCases),col=I('red'),lwd=1) + scale_x_date(limits = c(strtX,max(modelX$date)))+
          ggtitle('COVID-19 CZ growth rate last 5ys (vs doubling time N days)') + ylab('totalCases') + 
          scale_y_log10(limits=c(strtY,NA))+ geom_dl(aes(label = variable), method = list(dl.combine("last.points"), cex = 0.6)) + theme(legend.position='none')


```



### CZ data table

using USIZ data (apify)
```{r data table}

covidCZ$newTests <- c(0,diff(covidCZ$testedCases)) 
covidCZ$propIncreaseTests <- c(0,covidCZ$newTests[2:nrow(covidCZ)]/covidCZ$testedCases[1:(nrow(covidCZ)-1)])
covidCZ$propPosTested <- covidCZ$newCases/covidCZ$newTests


covidCZsm <- covidCZ[covidCZ$date>=as.Date('2020-02-22'),];

kable(covidCZsm)

```


### Austria vs CZ

AIM: estimate, by how many days the Czech Rep. is delayed behind other countries. First comparison: Austria

```{r comp-r}
covidCZjhu <- covid %>% filter(country=='Czech Republic') 
covidATjhu  <- covid %>% filter(country=='Austria') 

covidCZAT<-merge(covidCZjhu[,1:3],covidATjhu[,1:3],by='Date') %>% rename(casesCZ='cases.x',casesAT='cases.y') %>% select(Date,casesCZ,casesAT)


czDelay<-4

# covidCZAT %>% reshape2::melt(id='Date') %>% mutate(Date = ifelse(variable=='casesAT',Date,Date-days(czDelay))) %>%  ggplot(aes(x=Date,y=value,col=variable)) + geom_line(lwd=I(1))

 covidCZAT %>% mutate(casesATlag4d = lag(casesAT,n=czDelay)) %>% reshape2::melt(id='Date') %>% ggplot(aes(x=Date,y=value,col=variable)) + geom_line(lwd=I(1)) + geom_point() +xlim(as.Date('2020-02-26'),NA) + theme(legend.position = c(0.01,0.99),legend.justification = c(0,1)) + ggtitle(paste0('Total COVID-19 cases: Czech Rep. vs Austria (delayed by ',czDelay,' days)'))

 ggsave(paste0(figDir,'CzvsAt-4days-',dateStr,'.png'),width=20,height=15,units='cm',scale=.8)


# plot delayed data
qplot(covidCZAT$casesAT[1:(nrow(covidCZAT)-czDelay)],covidCZAT$casesCZ[(czDelay+1):nrow(covidCZAT)]) + xlab(paste0('casesAT delayed by ', czDelay,' days')) + ylab('casesCZ') +geom_abline(slope=1,intercept=0,col='orange') + ggtitle('Czech Republic vs delayed Austria: total cases')

# qplot(covidCZAT$Date, covidCZAT$casesAT[1:(nrow(covidCZAT)-czDelay)],geom='line') + geom_line(aes(x=covidCZAT$Date),y=covidCZAT$casesCZ[(czDelay+1):nrow(covidCZAT)]) + xlab(paste0('casesAT delayed by ', czDelay,' days')) + ylab('casesCZ') +geom_abline(slope=1,intercept=0,col='orange') + ggtitle('Czech Republic vs delayed Austria: total cases')



# covidCZAT %>% reshape2::melt(id='Date') %>% ggplot(aes(x=Date,y=value,col=variable)) + geom_line() + xlim(as.Date('2020-02-26'),NA) + ggtitle('COVID-19 CZ vs Austria')


```


### CZ vs others

The lag selected by minimizing the RMS between total cases in CZ and total cases in the comparison countries. This compares the initial segment only and we can thus see, which scenarios may follow...

```{r cz-delays}
covidCZjhu <- covid %>% filter(country=='Czech Republic') 


rms<-function(delayX,data){
  if(delayX>=0)
    sq<-data$x[(delayX+1):nrow(data)]-data$y[1:(nrow(data)-delayX)]
  else
    sq<-data$x[1:(nrow(data)+delayX)]-data$y[(-delayX+1):nrow(data)]
    
  return(sqrt(mean((sq)^2)))
  }
delayRange<--14:56



compCntrs <- c('Austria','Italy','Germany','Spain','France','Slovakia','Poland','China','South Korea')
compDelays<- rep(NA,length(compCntrs))
for(cntri in 1:length(compCntrs)){
  cntr<-compCntrs[cntri]
  # combine data
  covidComp<-covid %>% filter(country==cntr)
  covidCZComp<-merge(covidCZjhu[,1:3],covidComp[,1:3],by='Date')[,c('Date','cases.x','cases.y')] %>% rename(x='cases.x',y='cases.y')
  
  crit<-lapply(delayRange,rms,covidCZComp)
  compDelays[cntri] <- delayRange[which.min(crit)]
}

kable(t(rbind(compCntrs,compDelays)),col.names = c('Country','CZ delayed by (days)'))


# other ideas
# use r-squared (stupid idea)
# rsq<-function(czDelay,covidCZComp)summary(lm(log(1+covidCZComp$cases.x[(czDelay+1):length(covidCZComp$cases.x)])~log(1+covidCZComp$cases.y[1:(length(covidCZComp$cases.x)-czDelay)])))$r.squared
# lapply(1:10,rsq,covidCZComp)

# use RMS
# rmslg<-function(czDelay,covidCZComp)sqrt(mean((log(1+covidCZComp$cases.x[(czDelay+1):length(covidCZComp$cases.x)])-log(1+covidCZComp$cases.y[1:(length(covidCZComp$cases.x)-czDelay)]))^2))

```

Note: tested integer delays in the range `r min(delayRange)`, `r max(delayRange)` days


```{r plot-lagged}


compCntrs <- c(compCntrs,'Czech Republic')
compDelays<- c(compDelays,0)

covidLags<-covid %>% filter(country %in% compCntrs)
# lag the time series
for(cntr in compCntrs){
  covidLags$Date[covidLags$country==cntr] = covidLags$Date[covidLags$country==cntr] + days(compDelays[compCntrs==cntr])
  covidLags$country[covidLags$country==cntr] = paste0(cntr,': lag ',compDelays[compCntrs==cntr],' days')
}

pl<- covidLags[,c('country','Date','cases')] %>% ggplot(aes(x=Date,y=cases,col=country)) + geom_line(lwd=I(1)) + geom_point() +xlim(as.Date('2020-03-05'),NA)


pl+ geom_dl(aes(label = country), method = list(dl.combine("last.points"), cex = 0.6)) + theme(legend.position = 'none')


pl2 <- pl+xlim(Sys.Date()-days(7),Sys.Date()+days(7)) + ylim(0,6500) + theme(legend.position=c(0.05,0.95),legend.justification = c(0,1))


ref33 <- data.frame(Date=covidCZjhu$Date[which.max(covidCZjhu$Date)]+days(0:20), cases=covidCZjhu$cases[which.max(covidCZjhu$Date)]*1.33^c(0:20), country='33% daily') 

pl3<-pl2 + ggtitle('COVID19 trends: Czech Rep. vs lagged comparison countries') + ylab('total cases') + scale_x_date(breaks=max(covidCZjhu$Date)+days(seq(-14,13,7)),limits=c(max(covidCZjhu$Date)-days(14),max(covidCZjhu$Date)+days(13))) +geom_vline(xintercept =max(covidCZjhu$Date),col='orange' )+ geom_dl(aes(label = country), method = list(dl.combine("last.points"), cex = 0.6)) + geom_abline() + geom_line(data=ref33, linetype='dotted',lwd=1)


pl3 
ggsave(paste0(figDir,'CZvsLaggedOther',dateStr,'.png'),pl3,width=20,height=15,units='cm',scale=.8)
# pl2+facet_wrap(.~country)

pl4<-pl3 + scale_y_log10() + theme(legend.position = 'none')

pl4
ggsave(paste0(figDir,'CZvsLaggedOtherLOG',dateStr,'.png'),pl4,width=20,height=15,units='cm',scale=.8)
```


<!-- ### exploratory plots: JHU data -->
```{r plot-CZ-jhu}
# covidCZjhu <- covid %>% filter(country=='Czech Republic') 
# covidCZjhu[,1:6] %>% reshape2::melt(id=c('country','Date')) %>% ggplot(aes(x=Date,y=value,col=variable)) + geom_line() +colSc + ggtitle('COVID-19 Czech Republic')


```




### compare JHU and ÚZIS data

```{r compare-jhu-uzis}

combCZ <- merge(covidCZ,covidCZjhu,by.x='date',by.y='Date')
covidCZjhu <- covid %>% filter(country=='Czech Republic') 


combCZ[,c('date','totalCases','newCases','cases','incident')] %>% rename(totalUZIS='totalCases',newUZIS='newCases',totalJHU='cases',newJHU='incident') %>% reshape2::melt(id=c('date')) %>% ggplot(aes(x=date,y=value,col=variable)) + xlim(as.Date('2020-02-28'),NA) + geom_line() + ggtitle('CovidCZ: JHU and UZIS data comparison')

```

### Load public policy dates in CZ
```{r pub-pol}
 
 policyCZwiki<-loadPolicyCzWiki()
 policyCZzpl<-loadPolicyCZzakonyprolidi()

 kable(policyCZwiki,caption='Czech public policy changes (wikipedia)')
 
```



### Epidemiological modelling - TBD

TBD according to: https://rviews.rstudio.com/2020/03/05/covid-19-epidemiology-with-r/
And https://timchurches.github.io/blog/posts/2020-03-10-modelling-the-effects-of-public-health-interventions-on-covid-19-transmission-part-1/



```{r epi-cz}


```

## Store data locally for further ref

```{r save-todays-data}

 write.csv(covid,paste0(dataDir,'covid',dateStr,'.csv'))

 write.csv(covidCZ,paste0(dataDir,'covidCZ',dateStr,'.csv'))
 
 write.csv(policyCZwiki,paste0(dataDir,'policyCZwiki',dateStr,'.csv'))

 write.csv(policyCZzpl,paste0(dataDir,'policyCZzakonyprolidi',dateStr,'.csv'))

```

